---
title: "Canadian Crop Production for Brewing and Distilling"
subtitle: "A reproducible SQL analysis of production, yield, prices and exchange rates"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Business question

Canadian microbreweries and microdistilleries depend on grains whose availability and cost change across provinces and years. This notebook exposes the complete workflow: CSV ingestion, validation, SQLite construction, parameterized SQL, currency conversion, and visualization.

## Load packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(DBI)
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(RSQLite)
library(scales)
library(tibble)
```

## Read and validate four sources

```{r}
#| label: read-data
crop <- read.csv(here("data", "annual_crop.csv"), stringsAsFactors = FALSE)
daily_fx <- read.csv(here("data", "daily_fx.csv"), stringsAsFactors = FALSE)
monthly_fx <- read.csv(here("data", "monthly_fx.csv"), stringsAsFactors = FALSE)
farm_prices <- read.csv(
  here("data", "monthly_farm_prices.csv"),
  stringsAsFactors = FALSE
)

crop$YEAR <- as.Date(crop$YEAR)
crop$YEAR_NUMBER <- as.integer(format(crop$YEAR, "%Y"))
daily_fx$DATE <- as.Date(daily_fx$DATE)
monthly_fx$DATE <- as.Date(monthly_fx$DATE)
farm_prices$DATE <- as.Date(farm_prices$DATE)

required_crop_columns <- c(
  "CD_ID", "YEAR", "CROP_TYPE", "GEO", "SEEDED_AREA",
  "HARVESTED_AREA", "PRODUCTION", "AVG_YIELD"
)
required_fx_columns <- c("DFX_ID", "DATE", "FXUSDCAD")
required_price_columns <- c(
  "CD_ID", "DATE", "CROP_TYPE", "GEO", "PRICE_PRERMT"
)

stopifnot(
  all(required_crop_columns %in% names(crop)),
  all(required_fx_columns %in% names(daily_fx)),
  all(required_fx_columns %in% names(monthly_fx)),
  all(required_price_columns %in% names(farm_prices)),
  nrow(crop) > 0,
  nrow(daily_fx) > 0,
  nrow(monthly_fx) > 0,
  nrow(farm_prices) > 0,
  anyDuplicated(crop$CD_ID) == 0
)

tibble(
  table = c("crop_data", "daily_fx", "monthly_fx", "farm_prices"),
  records = c(
    nrow(crop),
    nrow(daily_fx),
    nrow(monthly_fx),
    nrow(farm_prices)
  )
) |>
  kable()
```

## Build SQLite during the render

The database is temporary, so the repository stores transparent CSV snapshots rather than a binary database.

```{r}
#| label: build-database
database_path <- tempfile(fileext = ".sqlite")
connection <- dbConnect(SQLite(), database_path)

dbWriteTable(connection, "crop_data", crop, overwrite = TRUE)
dbWriteTable(connection, "daily_fx", daily_fx, overwrite = TRUE)
dbWriteTable(connection, "monthly_fx", monthly_fx, overwrite = TRUE)
dbWriteTable(connection, "farm_prices", farm_prices, overwrite = TRUE)

dbListTables(connection)
```

## National production trends

The crop names are passed as bound parameters rather than pasted into SQL.

```{r}
#| label: production-query
production_sql <- "
  SELECT YEAR, CROP_TYPE, PRODUCTION
  FROM crop_data
  WHERE GEO = 'Canada'
    AND CROP_TYPE IN (?, ?, ?)
  ORDER BY YEAR, CROP_TYPE
"

production_trends <- dbGetQuery(
  connection,
  production_sql,
  params = list("Barley", "Rye", "Wheat")
)

production_trends$YEAR <- as.Date(
  production_trends$YEAR,
  origin = "1970-01-01"
)

latest_production <- production_trends |>
  filter(YEAR == max(YEAR)) |>
  slice_max(PRODUCTION, n = 1, with_ties = FALSE)
```

```{r}
#| label: production-chart
#| fig-alt: "Lines showing Canadian barley, rye, and wheat production over time."
ggplot(
  production_trends,
  aes(x = YEAR, y = PRODUCTION, color = CROP_TYPE)
) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_color_viridis_d(end = 0.85) +
  labs(
    title = "Canadian crop production over time",
    x = "Year",
    y = "Production (metric tonnes)",
    color = "Crop"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Highest yields in Saskatchewan

Geography, year, and result count are all bound parameters.

```{r}
#| label: yield-query
yield_sql <- "
  SELECT
    CROP_TYPE,
    ROUND(AVG(AVG_YIELD), 1) AS average_yield
  FROM crop_data
  WHERE GEO = ? AND YEAR_NUMBER = ?
  GROUP BY CROP_TYPE
  ORDER BY average_yield DESC
  LIMIT ?
"

top_yields <- dbGetQuery(
  connection,
  yield_sql,
  params = list("Saskatchewan", 2000L, 6L)
)

top_yields |>
  kable(col.names = c("Crop", "Average yield"))
```

## Canola prices in CAD and USD

The query joins monthly prices and exchange rates by date. Dividing CAD per tonne by CAD per USD produces USD per tonne.

```{r}
#| label: price-query
price_sql <- "
  SELECT
    p.DATE,
    p.PRICE_PRERMT AS price_cad_per_tonne,
    fx.FXUSDCAD,
    ROUND(p.PRICE_PRERMT / fx.FXUSDCAD, 2) AS price_usd_per_tonne
  FROM farm_prices p
  INNER JOIN monthly_fx fx ON p.DATE = fx.DATE
  WHERE p.CROP_TYPE = 'Canola' AND p.GEO = ?
  ORDER BY p.DATE DESC
  LIMIT ?
"

canola_prices <- dbGetQuery(
  connection,
  price_sql,
  params = list("Saskatchewan", 12L)
)

canola_prices$DATE <- as.Date(canola_prices$DATE, origin = "1970-01-01")

canola_prices |>
  slice_head(n = 6) |>
  kable(
    col.names = c("Date", "CAD / tonne", "CAD per USD", "USD / tonne"),
    digits = 2
  )
```

## Decision notes

- Production volume provides supply context, not immediately available inventory.
- Provincial detail matters for procurement decisions.
- Currency conversion is a dated estimate, not a transaction quote.
- Transportation, quality, and contracts remain outside this dataset.

```{r}
#| label: disconnect
dbDisconnect(connection)
```

## Conclusion

- Within the three national series, `r latest_production$CROP_TYPE[[1]]` has the largest production in `r format(latest_production$YEAR[[1]], "%Y")`, reaching `r format(latest_production$PRODUCTION[[1]], big.mark = ",", scientific = FALSE)` metric tonnes. This establishes the dominant volume signal in the latest shared year.
- The parameterized Saskatchewan query ranks `r top_yields$CROP_TYPE[[1]]` first in 2000 with an average yield of `r format(top_yields$average_yield[[1]], big.mark = ",")` in the source dataset's reported yield units.
- For `r format(canola_prices$DATE[[1]], "%B %Y")`, canola is `r sprintf("%.2f", canola_prices$price_cad_per_tonne[[1]])` CAD per tonne. At `r sprintf("%.4f", canola_prices$FXUSDCAD[[1]])` CAD per USD, the joined estimate is `r sprintf("%.2f", canola_prices$price_usd_per_tonne[[1]])` USD per tonne.
- These results support supply and currency monitoring, but not a purchasing decision by themselves: transport costs, quality grades, contract terms, and basis risk are absent from the relational model.

See the [Spanish version](es/index.qmd).
