---
title: "Canadian Crop Production for Brewing and Distilling"
subtitle: "A reproducible SQL analysis of production, yield, prices and exchange rates"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis.R"))
sources <- read_crop_sources()
connection <- build_crop_database(sources)
```

## Business question

Canadian microbreweries and microdistilleries depend on grains whose availability and cost change across provinces and years. This project combines agricultural production, farm prices, and CAD/USD exchange rates in SQLite to demonstrate transparent, reusable SQL analysis.

## Data quality and database design

The four snapshots contain annual crop observations, daily and monthly exchange rates, and monthly farm prices. Dates are parsed before loading, required columns are validated, and each render builds a temporary database instead of committing a binary database file.

```{r}
tibble::tibble(
  table = c("crop_data", "daily_fx", "monthly_fx", "farm_prices"),
  records = c(nrow(sources$crop), nrow(sources$daily_fx), nrow(sources$monthly_fx), nrow(sources$prices))
) |>
  knitr::kable()
```

## National production trends

```{r}
trends <- national_crop_trends(connection)
plot_crop_trends(trends, "en")
```

Production volume provides supply context, but it should not be interpreted as immediately available inventory for a specific buyer.

## Which crops had the highest yield?

The query below uses bound parameters for geography and year rather than constructing SQL from user-provided text.

```{r}
top_crop_yields(connection, "Saskatchewan", 2000, limit = 6) |>
  knitr::kable(col.names = c("Crop", "Average yield"))
```

## Canola prices in two currencies

Monthly farm prices are joined to monthly exchange rates on a shared date. Dividing the CAD price by CAD per USD converts the value to USD per metric tonne.

```{r}
canola_prices <- recent_canola_prices(connection)
canola_prices |>
  head(6) |>
  knitr::kable(
    col.names = c("Date", "CAD / tonne", "CAD per USD", "USD / tonne"),
    digits = 2
  )
```

## Decision notes

- Compare long-run production with recent farm prices before choosing a grain.
- Preserve geographic detail; national totals cannot describe provincial procurement conditions.
- Treat exchange-rate conversion as a dated estimate, not a quoted transaction price.
- Add transportation, quality specifications, and supplier contracts before making purchasing decisions.

## Conclusion

SQLite makes the analytical transformations inspectable and repeatable. Parameterized queries, explicit units, and temporary database construction improve the original exercise while keeping its business focus.

```{r}
#| include: false
DBI::dbDisconnect(connection)
```
